Supplementary Materials for CBEA: Competitive Balances for Taxonomic Enrichment Analysis

### 2022-02-15 supplemental.tex
Updates from Overleaf

\subsubsection{Prediction analysis}
Fig~\ref{fig:s5} shows results for simulation studies as detailed in the \nameref{methods} section. Panel A presents results for a regression learning task with a continuous outcome while panel B presents results for a classification task with a binary outcome. As expected, performance across all assessed methods increased with a higher signal-to-noise ratio. Both CLR and CBEA approaches outperformed both GSVA and ssGSEA across all simulation conditions and learning tasks. This is because both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity, while CBEA and CLR did not experience a similar level of impact. As such, performance gap widens with increasing correlation and sparsity. Interestingly, this difference in performance is not as pronounced under high levels of effect saturation (across both learning tasks), suggesting that when there is a high number of sets contributing to an effect, model choice might not be as important.

In this analysis, CBEA unfortunately did not outperform the CLR approach, which is standard practice within the microbiome literature \cite{gloor2017}. This difference in performance is more notable in regression learning tasks compared to classification, and at lower levels of effect saturation. However, the degree of separation between the two approaches is not as dramatic as between GSVA/ssGSEA and CBEA/CLR. Moreover, the performance gap decreases with increasing effect signal-to-noise ratio and sparsity. Additionally, we did not observe any performance difference between the different variations of CBEA.

\begin{figure}[!h]
    \centering
    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
    \caption{Predictive performance of a naive random forest model trained on CBEA, ssGSEA, GSVA generated scores as well as the standard CLR approach on simulated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio. Panel \textbf{(A)} presents performance on a regression task using predictive R-squared as the evaluation measure. Panel \textbf{(B)} presents performance on a classification task with AUROC as the evaluation measure. CBEA approaches outperformed GSVA and ssGSEA across all simulation conditions but not the CLR approach.}
    \label{fig:s5}
\end{figure}

\section{Performance evaluation} 
We implemented CBEA in the package \emph{CBEA} on GitHub (\url{https://www.github.com/qpmnguyen/CBEA}). We evaluated computational time using the \emph{bench} package in R. We applied CBEA to a standard data set generated using our simulation model consisting of 40 sets (of size 20 each) and 500 samples. Benchmark was performed on a single core using a node on the computing cluster (Specifications: Intel Xeon E5-2690 (2.6GHz) with 4GB of RAM).

@@

In order to reduce runtime, users can attempt the following: Since CBEA fits parametric distributions over permuted values of all samples within a data set (i.e. for $N = 100$ and $10$ permutations, the fitting procedure will attempt to estimate parameters from a vector of size $1000$ per set), if the sample size is high users can reduce the number of permutations. Additionally, CBEA also implements a procedure to parallelize computation across sets, which might be applicable to situations where there are a lot of sets. Finally, a lot of CBEA approaches work well without parametric fit, so users can use the non-parametric approaches.

>>>>

\subsubsection{Prediction analysis}

\begin{figure}[!h]
    \centering
    \includegraphics[width = \linewidth]{figures/sim_pred_combined.png}
    \caption{\textbf{Simulation results for predictive performance evaluation for CBEA}. Predictive performance of a random forest model (with no hyperparameter tuning) trained on set-based features as inputs. Methods to generate these features include CBEA, ssGSEA, GSVA, and the CLR transformation applied on sum-aggregated sets. Simulation data was generated across different levels of data sparsity, inter-taxa correlation, effect saturation, and signal-to-noise ratio. Panel \textbf{(A)} presents performance on a regression task using RMSE (root mean squared error) as the evaluation measure. Panel \textbf{(B)} presents performance on a classification task with AUROC as the evaluation measure}
    \label{fig:s5}
\end{figure}

Fig~\ref{fig:s5} shows results for simulation studies as detailed in the \nameref{methods} section. Panel A presents results for a regression learning task with a continuous outcome. We observed no difference in performance values across different evaluation models. Overall, prediction error did not change across sparsity levels, and decreases with increasing signal-to-noise ratio (SNR). However, at higher correlation levels, the pattern was more erratic. For example, when $\rho$ is set at $0.5$, higher SNR decreases performance only at low sparsity $p = 0.2$, but had the expected pattern medium sparsity $p = 0.4$, where higher SNR correlates with improved performance. This effect seems to be specific to low effect saturation scenarios (only a limited number of taxa sets are associated with the outcome). Interestingly, higher sparsity levels produces better performance.

Panel B represent results for a classification task with a binary outcome with AUROC as the evaluation criteria. Here we observed similar results as that of our real data evaluations, where using CLR transfromed data produces more predictive models across all scenarios. Conversely, GSVA and ssGSEA were consistently under performing when compared to CBEA and CLR. Interestingly, the degree of difference varies across sparsity and inter-taxa correlations. We noticed that increasing sparsity and correlation decreases the gap in performance between CBEA and CLR, while increasing the gap in performance between CBEA/CLR and GSVA/ssGSEA. As such, we can hypothesize that both GSVA and ssGSEA are more sensitive to the degree of inter-taxa correlation and sparsity. Finally, effect saturation did not change model rankings, but did decrease overall performance.

\section{Runtime} 
We implemented CBEA in the package \emph{CBEA} on GitHub (\url{https://www.github.com/qpmnguyen/CBEA}). We evaluated computational time using the \emph{bench} package in R. We applied CBEA to a standard data set generated using our simulation model consisting of 40 sets (of size 20 each) and 500 samples. Benchmark was performed on a single core using a node on the computing cluster (Specifications: Intel Xeon E5-2690 (2.6GHz) with 4GB of RAM).

@@

In order to reduce runtime, users can attempt the following: Since CBEA fits parametric distributions over permuted values of all samples within a data set (i.e. for $N = 100$ and $10$ permutations, the fitting procedure will attempt to estimate parameters from a vector of size $1000$ for each set), if the sample size is high users can reduce the number of permutations. Additionally, CBEA also implements a procedure to parallelize computation across sets, which might be applicable to situations where there are a lot of sets to evaluate. Finally, a lot of CBEA approaches work well without parametric fit, so users can use the non-parametric approaches like the permutation test or using raw CBEA scores.
